Who Stays Single? Exploring the Factors Behind Non-Marriage in the United States in 2023

### 2024-12-04 paper/paper.qmd
Final commit

---
title: "Who Stays Single? Exploring the Factors Behind Non-Marriage in the United States in 2023"
subtitle: "Low Income and Education Levels Correlate with Higher Propensity to Remain Unmarried"
author: 
  - Xizi Sun
thanks: "Code and data are available at: https://github.com/LilianS77/US_Marriage."
date: today
date-format: long
abstract: "This study examines the socio-demographic determinants of non-marriage in the United States in 2023 using logistic regression analysis on individual-level data. The findings highlight that individuals with lower income and education levels, particularly below a bachelor’s degree, are significantly more likely to remain unmarried. Additionally, gender and race also play important roles, with men and certain racial groups, such as African Americans, showing higher unmarried rates. These results underscore structural inequities and socio-economic barriers shaping marital outcomes, offering insights into broader societal patterns and potential areas for intervention."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

The structure of this paper is as follows: @sec-data explains the dataset, variable selection, and measurement strategies employed to study marital status. @sec-model discusses the logistic regression methodology, including model rationale, prior specifications, and validation processes. @sec-results presents the key findings with supporting visualizations, emphasizing the roles of education, income, race, gender, and age in predicting non-marriage. Finally, @sec-discussion discusses the implications of the findings, acknowledges study limitations, and outlines potential areas for future research.

@@

## Measurement

@@

## Outcome Variables
The primary outcome variable for this study is **Marital Status**, which categorizes individuals based on their marital state. The proportion of marital status categories is displayed in @fig-proportion-marriage. This variable allows for a comparison between individuals who have never married (`Not_Married`) and those who have (`Married`). For this study,
**Not_Married**: Includes individuals who have never been married.
**Married**: Includes individuals who are married, as well as those who are divorced, widowed, or separated.

@@

## Determinants of Non-Marriage and Societal Implications

This study investigates the socio-demographic factors influencing marital status, focusing on the probability of remaining unmarried in the United States. Through logistic regression analysis applied to individual-level microdata, significant predictors such as education, income, race, gender, and age were identified.

Higher education levels were strongly associated with lower non-marriage rates. For instance, @fig-edu-marriage shows that only 16% of individuals with education above a bachelor’s degree were unmarried, compared to 32% for those with "Some College" education. Logistic regression coefficients further underscore this trend, with individuals holding higher degrees being substantially less likely to remain unmarried (e.g., the coefficient for "Below High School" = 0.48). Education thus acts as a stabilizing factor, equipping individuals with economic resources and social networks conducive to marriage.

Similarly, income plays a complementary but less pronounced role. @fig-Income-Unmarried illustrates that unmarried individuals tend to cluster in lower income brackets, with the majority earning less than $40,000 annually. This highlights the persistent importance of financial stability in marital decisions, although its impact appears secondary to education.

The results also emphasize significant disparities across racial and gender lines. African Americans exhibited the highest non-marriage rates (43%), as shown in @fig-race-marriage, significantly surpassing those of Asian Americans (30%) and White individuals (24%). These patterns likely reflect structural inequalities, such as higher unemployment rates and barriers to wealth accumulation among African Americans. Gender differences were also evident, with men showing a slightly higher likelihood of remaining unmarried compared to women (30% versus 26%, respectively, @fig-gender-marriage). This aligns with cultural norms where women are more closely tied to family roles and experience greater societal pressure to marry.

These findings reinforce how individual socio-demographic traits intersect with societal structures to shape marital outcomes. They also highlight systemic inequities that warrant targeted interventions.

## Analysis of Education, Income, and Marriage Patterns

Education emerged as the most significant predictor in this analysis. Higher education levels correlate strongly with lower non-marriage rates, as depicted in @fig-edu-marriage and @fig-coefficient. These findings align with theories suggesting that education enhances economic stability, expands social capital, and provides greater agency in life choices, all of which promote marriage. However, as @fig-Income-Unmarried indicates, financial stability remains an essential but secondary factor in marital decisions.

The interplay between education and income underscores broader socio-economic disparities. For example, individuals with lower education levels not only face limited economic opportunities but are also more likely to remain unmarried, perpetuating cycles of inequality in family formation.

## Gender and Race in Marital Trends

The analysis shows gender and racial disparities. As shown in @fig-gender-marriage, men exhibit higher non-marriage rates compared to women, likely reflecting cultural expectations and differing societal pressures. Racial disparities, however, are even more striking. African Americans, for instance, show the highest rates of non-marriage, suggesting systemic inequities such as income disparities and historical barriers to wealth accumulation. By contrast, Asian Americans, with the lowest non-marriage rates, may benefit from cultural norms emphasizing family cohesion and marriage stability.

These patterns illustrate how cultural norms and structural inequities converge to shape marital outcomes. Addressing these disparities requires policies that promote economic equity and social inclusivity.

@@

While this study provides useful insights into the determinants of non-marriage, it is not without limitations. The cross-sectional nature of the data restricts causal inferences, making it difficult to ascertain how variables such as education and income evolve over time to influence marital decisions. Longitudinal data would allow for a deeper exploration of these dynamics.

>>>>

---
title: "Who Stays Single? Exploring the Factors Behind Non-Marriage in the United States in 2023"
subtitle: "Low Income and Education Levels Correlate with Higher Propensity to Remain Unmarried"
author: 
  - Xizi Sun
thanks: "Code and data are available at: https://github.com/LilianS77/US_Marriage."
date: today
date-format: long
abstract: "This study examines the socio-demographic factors influencing non-marriage in the United States in 2023 through logistic regression analysis of individual-level data. The findings show that individuals with lower income and education levels, particularly those without a bachelor’s degree, are more likely to remain unmarried. Additionally, gender and race significantly influence outcomes, with men and certain racial groups, such as African Americans, experiencing higher rates of non-marriage. These results highlight structural inequities and socio-economic barriers affecting marital patterns and identify areas for potential policy action."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

The structure of this paper is as follows: @sec-data explains the dataset, variable selection, and measurement strategies employed to study marital status. @sec-model discusses the logistic regression methodology. @sec-results presents the key findings with supporting visualizations, emphasizing the roles of education, income, race, gender, and age in predicting non-marriage. Finally, @sec-discussion discusses the implications of the findings, acknowledges study limitations, and outlines potential areas for future research.

@@

## Data Source and Measurement

The data employed in this research was obtained from IPUMS USA (Integrated Public Use Microdata Series) [@ipumsusa]. This dataset contains individual-level microdata, where each record represents a distinct individual. The attributes of these individuals are numerically coded and span a wide array of demographic, economic, and social variables.

@@

## Outcome Variables
The primary outcome variable for this study is **Marital Status**, which categorizes individuals based on their marital state. The proportion of marital status categories is displayed in @fig-proportion-marriage. This variable allows for a comparison between individuals who have never married (`Not_Married`) and those who have (`Married`). For this study,

1. **Not_Married**: Includes individuals who have never been married.
2. **Married**: Includes individuals who are married, as well as those who are divorced, widowed, or separated.

@@

## Determinants of Non-Marriage

This study identifies the socio-demographic factors influencing marital status in the United States, focusing on the likelihood of remaining unmarried. Through logistic regression analysis, the roles of education, income, race, gender, and age were examined.

### The Role of Education

Education was found to be the most significant factor associated with marital status. Individuals with higher educational attainment, such as those holding graduate degrees, had the lowest rates of non-marriage (16%), while those with “Some College” education exhibited a 32% likelihood of being unmarried. The analysis highlights how education contributes to economic stability and social networks, creating conditions that favor marriage.  Aughinbaugh further observed that individuals with higher education are less likely to divorce and more likely to maintain stable marriages [@citeMarriagePatterns].

### Income as a Supporting Factor

Income played a secondary but meaningful role in marital decisions. As shown in @fig-Income-Unmarried, the majority of unmarried individuals were concentrated in lower income brackets, earning less than $40,000 annually. This indicates that financial stability can influence decisions about marriage. Stable economic conditions are likely to increase individuals’ willingness to marry and form families. These findings highlight the relationship between economic factors and broader marriage trends.

## Race and Gender Differences

### Racial Disparities

Racial differences were pronounced in the analysis. African Americans had the highest non-marriage rates (43%), compared to 30% for Asian Americans and 24% for White individuals. These disparities may reflect systemic inequities, such as limited access to wealth and higher unemployment rates [@citeAltonjiMarriage]. In contrast, cultural values that emphasize family cohesion may contribute to the lower rates among Asian Americans.

### Gender Disparities

Gender also shaped marital patterns, with men showing a slightly higher likelihood of being unmarried than women (30% vs. 26%). This trend aligns with cultural expectations, where women often face greater societal pressure to marry and assume family roles [@citeMarriagePatterns]. These findings suggest that cultural norms significantly affect marital outcomes.

## The Role of Age

Age patterns in marital status were evident in @fig-Age-Unmarried, with younger individuals, particularly those in their 20s and early 30s, being more likely to remain unmarried. This likelihood decreased with age, as individuals tend to enter marriage at more mature life stages when personal and financial readiness align. This trend is consistent with previous research, which shows that delayed marriage has become a common demographic pattern in modern society [@citeMarriagePatterns].

@@

While this study provides detailed analysis of the determinants of non-marriage, it is not without limitations. The cross-sectional nature of the data restricts causal inferences, making it difficult to ascertain how variables such as education and income evolve over time to influence marital decisions. Longitudinal data would allow for a deeper exploration of these dynamics.
